---
title: "FIFA World Cup Data Analysis"
author: "Nguyễn Đào Trường Vy"
date: "07-16-2026"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

library(tidyverse)
```

# Student Information

**Student Name:** Nguyễn Đào Trường Vy

**Student Number:** 22202382

**Subject Code:** COMP1013

**Subject Name:** Analytics Programming

**GitHub Repository:**  
https://github.com/yvette195/COMP1013-FIFA-WorldCup-Analysis

# Declaration

By including this statement, we the authors of this work, verify that:

- We hold a copy of this assignment that we can produce if the original is lost or damaged.

- We hereby certify that no part of this assignment/product has been copied from any other student’s work or from any other source except where due acknowledgement is made in the assignment.

- No part of this assignment/product has been written/produced for us by another person except where such collaboration has been authorised by the subject lecturer/tutor concerned.

- We are aware that this work may be reproduced and submitted to plagiarism detection software programs for the purpose of detecting possible plagiarism, which may retain a copy on its database for future plagiarism checking.

- We hereby certify that we have read and understand what the School of Computing, Engineering and Mathematics defines as minor and substantial breaches of misconduct as outlined in the learning guide for this unit.

\newpage

# Question 1: Data Inspection, Cleaning and Home Penalty Distribution

## Rationale

The four datasets were joined so that match, team, stadium and tournament information could be analysed together. Missing values `"?"` were changed to `NA`. The variables `Result`, `Stage`, `Country` and `ExtraTime` were changed to factors because they contain categories. `HomePenalty` and `AwayPenalty` were changed to numeric variables because they contain numbers of penalty goals.

A bar chart was choosen because `HomePenalty` contains distinct integer values. The height of each bar shows how many penalty shootouts had penalty scores for each home team.

## Load and Join the Data

```{r, question-1-load}
# Read the data CSV files.
matches <- read_csv("../data/Matches.csv")
stadiums <- read.csv("../data/Stadiums.csv")
teams <- read.csv("../data/Teams.csv")
tournaments <- read.csv("../data/Tournaments.csv")

# Change every "?" value to NA.
matches[matches == "?"] <- NA
stadiums[stadiums == "?"] <- NA
teams[teams == "?"] <- NA
tournaments[tournaments == "?"] <- NA
```

```{r, question-1-join}
# Join team information for the home team and away team.
df <- matches %>%
  left_join(
    teams %>%
      rename(
        HomeTeamID = TeamID,
        HomeTeamName = TeamName,
        HomeTeamCode = TeamCode
      ),
    by = "HomeTeamID"
  ) %>%
  left_join(
    teams %>%
      rename(
        AwayTeamID = TeamID,
        AwayTeamName = TeamName,
        AwayTeamCode = TeamCode
      ),
    by = "AwayTeamID"
  ) %>%
  left_join(stadiums, by = "StadiumID") %>%
  left_join(tournaments, by = "TournamentID")
```

## Inspect and Clean the Data

```{r, question-1-inspect}
# Display the first six rows.
head(df)

# Check the number of rows and columns.
dim(df)

# Check the variable types.
str(df)

# Display a summary of the main variables.
summary(
  df[c(
    "Result",
    "Stage",
    "Country",
    "ExtraTime",
    "PenaltyShootout",
    "HomePenalty",
    "AwayPenalty"
  )]
)
```

```{r question-1-clean}
# Convert categorical variables to factors.
df$Result <- as.factor(df$Result)
df$Stage <- as.factor(df$Stage)
df$Country <- as.factor(df$Country)
df$ExtraTime <- as.factor(df$ExtraTime)

# Convert penalty variables to numeric.
df$HomePenalty <- as.numeric(df$HomePenalty)
df$AwayPenalty <- as.numeric(df$AwayPenalty)

# Replace missing penalty values with zero.
df$HomePenalty[is.na(df$HomePenalty)] <- 0
df$AwayPenalty[is.na(df$AwayPenalty)] <- 0
```

## Code Testing

```{r question-1-testing}
# The number of rows should be unchanged after joining.
nrow(matches)
nrow(df)

# Check the classes after conversion.
class(df$Result)
class(df$Stage)
class(df$Country)
class(df$ExtraTime)
class(df$HomePenalty)
class(df$AwayPenalty)

# Check that no missing penalty values remain.
sum(is.na(df$HomePenalty))
sum(is.na(df$AwayPenalty))
```

The two row counts should be equal. The four categorical variables should return `"factor"`, the two penalty variables should return `"numeric"`, and both missing-value checks should return zero.

## Results and Interpretation

```{r question-1-results}
# Keep only matches involving a penalty shootout.
penalty_matches <- df %>%
  filter(PenaltyShootout == 1)

# Create a frequency table for HomePenalty.
penalty_table <- table(penalty_matches$HomePenalty)
penalty_table

# Find the most common HomePenalty value.
most_common_penalty <- names(penalty_table)[which.max(penalty_table)]
most_common_count <- max(penalty_table)
```

```{r question-1-plot, fig.width=8, fig.height=5}
ggplot(penalty_matches, aes(x = factor(HomePenalty))) +
  geom_bar() +
  labs(
    title = "Distribution of Home-Team Penalty Goals",
    x = "Home-team penalty goals",
    y = "Number of matches"
  ) +
  theme_minimal()
```

There were `r nrow(penalty_matches)` matches involving a penalty shootout. The most common value of `HomePenalty` was `r most_common_penalty`, and it occurred in `r most_common_count` matches. The taller bars show the penalty scores that occurred most often, while shorter bars represent less common scores.

\newpage

# Question 2: Distribution of Home Team Scores

## Rationale

`HomeTeamScore` is a numerical variable, so histograms were used to show its distribution. A bin width of one was selected because football scores are whole numbers. The distributions were compared first by `Stage`, then by `TournamentName`, and finally by the three required `AwayTeamScore` groups.

The `aggregate()` function was used to calculate simple group means and medians. This provides numerical information to support the patterns shown in the histograms.

## Distribution by Competition Stage

```{r question-2-stage-summary}
# Calculate the mean and median HomeTeamScore for each stage.
stage_mean <- aggregate(
  HomeTeamScore ~ Stage,
  mean,
  data = df
)

stage_median <- aggregate(
  HomeTeamScore ~ Stage,
  median,
  data = df
)

stage_mean
stage_median
```

```{r question-2-stage-plot, fig.width=9, fig.height=7}
ggplot(df, aes(x = HomeTeamScore)) +
  geom_histogram(binwidth = 1) +
  facet_wrap(~ Stage) +
  labs(
    title = "Home-Team Scores by Competition Stage",
    x = "Home-team goals",
    y = "Number of matches"
  ) +
  theme_minimal()
```

## Distribution by Tournament Name

```{r question-2-tournament-summary}
# Calculate the mean and median HomeTeamScore for each tournament.
tournament_mean <- aggregate(
  HomeTeamScore ~ TournamentName,
  mean,
  data = df
)

tournament_median <- aggregate(
  HomeTeamScore ~ TournamentName,
  median,
  data = df
)

tournament_mean
tournament_median
```

```{r question-2-tournament-plot, fig.width=10, fig.height=12}
ggplot(df, aes(x = HomeTeamScore)) +
  geom_histogram(binwidth = 1) +
  facet_wrap(~ TournamentName) +
  labs(
    title = "Home-Team Scores by Tournament",
    x = "Home-team goals",
    y = "Number of matches"
  ) +
  theme_minimal()
```

## Distribution by Away-Team Score Group

```{r question-2-away-groups}
# Create an empty variable for the away-team score groups.
df$AwayScoreGroup <- NA

# Assign each match to one of the three required groups.
df$AwayScoreGroup[df$AwayTeamScore <= 1] <- "0-1 goals"

df$AwayScoreGroup[
  df$AwayTeamScore >= 2 &
    df$AwayTeamScore <= 3
] <- "2-3 goals"

df$AwayScoreGroup[df$AwayTeamScore >= 4] <- "4 or more goals"

# Convert the new variable to a factor.
df$AwayScoreGroup <- as.factor(df$AwayScoreGroup)

# Check the number of matches in each group.
table(df$AwayScoreGroup)

# Calculate the mean and median HomeTeamScore for each group.
away_group_mean <- aggregate(
  HomeTeamScore ~ AwayScoreGroup,
  mean,
  data = df
)

away_group_median <- aggregate(
  HomeTeamScore ~ AwayScoreGroup,
  median,
  data = df
)

away_group_mean
away_group_median
```

```{r question-2-away-plot, fig.width=9, fig.height=5}
ggplot(df, aes(x = HomeTeamScore)) +
  geom_histogram(binwidth = 1) +
  facet_wrap(~ AwayScoreGroup) +
  labs(
    title = "Home-Team Scores by Away-Team Score Group",
    x = "Home-team goals",
    y = "Number of matches"
  ) +
  theme_minimal()
```

## Code Testing

```{r question-2-testing}
# Check the score variable types.
class(df$HomeTeamScore)
class(df$AwayTeamScore)

# Check for missing score values.
sum(is.na(df$HomeTeamScore))
sum(is.na(df$AwayTeamScore))

# Check the new group variable.
class(df$AwayScoreGroup)
table(df$AwayScoreGroup)
sum(is.na(df$AwayScoreGroup))

# Check that all matches are included in the three groups.
sum(table(df$AwayScoreGroup))
nrow(df)
```

`HomeTeamScore` and `AwayTeamScore` should be numeric variables; R may display their class as `integer` because football goals are whole numbers. The missing-value checks should return zero. `AwayScoreGroup` should be a factor, and the total number shown by its frequency table should equal the total number of rows in `df`.

## Results and Interpretation

```{r question-2-interpretation-values, include=FALSE}
# Find the stage with the highest and lowest mean HomeTeamScore.
highest_stage_row <- which.max(stage_mean$HomeTeamScore)
lowest_stage_row <- which.min(stage_mean$HomeTeamScore)

highest_stage <- stage_mean$Stage[highest_stage_row]
highest_stage_mean <- stage_mean$HomeTeamScore[highest_stage_row]

lowest_stage <- stage_mean$Stage[lowest_stage_row]
lowest_stage_mean <- stage_mean$HomeTeamScore[lowest_stage_row]

# Find the tournament with the highest and lowest mean HomeTeamScore.
highest_tournament_row <- which.max(tournament_mean$HomeTeamScore)
lowest_tournament_row <- which.min(tournament_mean$HomeTeamScore)

highest_tournament <- tournament_mean$TournamentName[highest_tournament_row]
highest_tournament_mean <- tournament_mean$HomeTeamScore[highest_tournament_row]

lowest_tournament <- tournament_mean$TournamentName[lowest_tournament_row]
lowest_tournament_mean <- tournament_mean$HomeTeamScore[lowest_tournament_row]

# Find the away-score group with the highest mean HomeTeamScore.
highest_group_row <- which.max(away_group_mean$HomeTeamScore)
highest_group <- away_group_mean$AwayScoreGroup[highest_group_row]
highest_group_mean <- away_group_mean$HomeTeamScore[highest_group_row]
```

The stage with the highest average home-team score was **`r highest_stage`**, with an average of `r round(highest_stage_mean, 2)` goals. The lowest average occurred in **`r lowest_stage`**, with `r round(lowest_stage_mean, 2)` goals. The stage histograms also show whether the scores were concentrated at low values or spread across a wider range.

The tournament with the highest average home-team score was **`r highest_tournament`**, with `r round(highest_tournament_mean, 2)` goals. The lowest tournament average occurred in **`r lowest_tournament`**, with `r round(lowest_tournament_mean, 2)` goals. The tournament histograms make it possible to compare how the score distributions changed between World Cup editions.

Among the three away-team score groups, **`r highest_group`** had the highest average home-team score at `r round(highest_group_mean, 2)` goals. This result describes a relationship in the dataset, but it does not prove that the away-team score caused the home-team score.

\newpage

# Question 3: Penalty Shootout Participation

## Rationale

Only matches where `PenaltyShootout` equals 1 were selected because this question focuses on matches decided by a penalty shootout. The matches were grouped separately by home team and away team because the assignment requires two different rankings.

The number of matches for each team was counted using `group_by()` and `summarise()`. The results were sorted from highest to lowest, and `head()` was used to select the first five teams.

## Filter Penalty Shootout Matches

```{r question-3-filter}
# Keep only matches that involved a penalty shootout.
penalty_matches <- df %>%
  filter(PenaltyShootout == 1)

# Check the number of penalty-shootout matches.
nrow(penalty_matches)

# Check that the filtered data contains only value 1.
table(penalty_matches$PenaltyShootout)
```

## Top Five Home Teams

```{r question-3-home-teams}
# Group the penalty matches by home-team name and code.
# Count the number of matches for each home team.
home_penalty_count <- penalty_matches %>%
  group_by(HomeTeamName, HomeTeamCode) %>%
  summarise(NumberOfMatches = n())

# Sort the teams from the highest count to the lowest count.
home_penalty_count <- home_penalty_count %>%
  arrange(desc(NumberOfMatches))

# Select the first five teams.
top5_home <- head(home_penalty_count, 5)

# Change the column names to match the assignment requirement.
names(top5_home) <- c(
  "TeamName",
  "TeamCode",
  "NumberOfMatches"
)

# Display the top five home teams as a table.
top5_home
```

## Top Five Away Teams

```{r question-3-away-teams}
# Group the penalty matches by away-team name and code.
# Count the number of matches for each away team.
away_penalty_count <- penalty_matches %>%
  group_by(AwayTeamName, AwayTeamCode) %>%
  summarise(NumberOfMatches = n())

# Sort the teams from the highest count to the lowest count.
away_penalty_count <- away_penalty_count %>%
  arrange(desc(NumberOfMatches))

# Select the first five teams.
top5_away <- head(away_penalty_count, 5)

# Change the column names to match the assignment requirement.
names(top5_away) <- c(
  "TeamName",
  "TeamCode",
  "NumberOfMatches"
)

# Display the top five away teams as a table.
top5_away
```

## Code Testing

```{r question-3-testing}
# Check that all filtered matches involved a penalty shootout.
table(penalty_matches$PenaltyShootout)

# Both top-five tables should contain five rows.
nrow(top5_home)
nrow(top5_away)

# Check that both tables have the required column names.
names(top5_home)
names(top5_away)

# Check for missing values in team names and codes.
sum(is.na(top5_home$TeamName))
sum(is.na(top5_home$TeamCode))

sum(is.na(top5_away$TeamName))
sum(is.na(top5_away$TeamCode))

# Check that the match counts are sorted from high to low.
top5_home$NumberOfMatches
top5_away$NumberOfMatches
```

The `PenaltyShootout` table should contain only the value 1. Both top-five tables should contain five rows and the columns `TeamName`, `TeamCode`, and `NumberOfMatches`. The missing-value checks should return zero. The match counts should be displayed from the highest value to the lowest value.

## Results and Interpretation

```{r question-3-comparison, include=FALSE}
# Find teams that appear in both top-five tables.
common_teams <- top5_home$TeamName[
  top5_home$TeamName %in% top5_away$TeamName
]

# Count the number of teams appearing in both lists.
number_common <- length(common_teams)

# Check whether the two rankings have the same teams in the same order.
same_ranking <- identical(
  top5_home$TeamName,
  top5_away$TeamName
)

# Store the leading home and away teams for the discussion.
leading_home_team <- top5_home$TeamName[1]
leading_home_count <- top5_home$NumberOfMatches[1]

leading_away_team <- top5_away$TeamName[1]
leading_away_count <- top5_away$NumberOfMatches[1]
```

The national team ranked first in the home-team table was **`r leading_home_team`**, which participated in `r leading_home_count` penalty-shootout matches as the home team. The national team ranked first in the away-team table was **`r leading_away_team`**, which participated in `r leading_away_count` matches as the away team.

There were `r number_common` teams appearing in both top-five lists: **`r paste(common_teams, collapse = ", ")`**. The rankings were **`r ifelse(same_ranking, "the same", "different")`** when the team names and their order were compared.

The difference between the two rankings may occur because teams can be recorded as the home team in some matches and the away team in other matches. These labels do not necessarily mean that the match was played in the team’s own country.

\newpage

# Question 4: Factors Influencing Match Outcomes

## Selected Factors and Rationale

The two selected factors were `Stage` and `ExtraTime`.

`Stage` was selected because match conditions may differ between group-stage matches and knockout-stage matches. For example, teams may use different strategies in finals, semi-finals, or group-stage matches.

`ExtraTime` was selected because matches that continue beyond regular time may have a different distribution of results compared with matches that finish during regular time.

Frequency tables were used to show the number of home-team wins, away-team wins, and draws. Percentage tables were also created because the numbers of matches in the different groups are not equal. A proportional stacked bar chart was selected because it allows the three result categories to be compared within each group.

## Overall Match Results

```{r question-4-overall-results}
# Count the number of matches in each Result category.
overall_result_table <- table(df$Result)

overall_result_table

# Find the most common match result.
most_common_result <- names(overall_result_table)[
  which.max(overall_result_table)
]

most_common_result_count <- max(overall_result_table)
```

## Factor 1: Competition Stage

```{r question-4-stage-table}
# Create a table of Stage and Result.
stage_result_count <- table(
  df$Stage,
  df$Result
)

stage_result_count

# Convert the counts into row percentages.
# margin = 1 calculates percentages within each Stage.
stage_result_percent <- prop.table(
  stage_result_count,
  margin = 1
) * 100

# Round the percentages to one decimal place.
stage_result_percent <- round(
  stage_result_percent,
  1
)

stage_result_percent
```

```{r question-4-stage-plot, fig.width=9, fig.height=6}
# Create a proportional bar chart of Result for each Stage.
ggplot(df, aes(x = Stage, fill = Result)) +
  geom_bar(position = "fill") +
  labs(
    title = "Match Results by Competition Stage",
    x = "Competition stage",
    y = "Proportion of matches",
    fill = "Match result"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```

## Factor 2: Extra Time

```{r question-4-extra-time-table}
# Create a table of ExtraTime and Result.
extra_time_result_count <- table(
  df$ExtraTime,
  df$Result
)

extra_time_result_count

# Convert the counts into row percentages.
# margin = 1 calculates percentages within each ExtraTime group.
extra_time_result_percent <- prop.table(
  extra_time_result_count,
  margin = 1
) * 100

# Round the percentages to one decimal place.
extra_time_result_percent <- round(
  extra_time_result_percent,
  1
)

extra_time_result_percent
```

```{r question-4-extra-time-plot, fig.width=8, fig.height=5}
# Create a proportional bar chart of Result by ExtraTime.
ggplot(df, aes(x = ExtraTime, fill = Result)) +
  geom_bar(position = "fill") +
  labs(
    title = "Match Results by Extra-Time Category",
    x = "Extra-time category",
    y = "Proportion of matches",
    fill = "Match result"
  ) +
  theme_minimal()
```

## Code Testing

```{r question-4-testing}
# Check that the three variables are factors.
class(df$Result)
class(df$Stage)
class(df$ExtraTime)

# Display the categories in each variable.
levels(df$Result)
levels(df$Stage)
levels(df$ExtraTime)

# Check for missing values.
sum(is.na(df$Result))
sum(is.na(df$Stage))
sum(is.na(df$ExtraTime))

# The total number in each table should equal the number of rows in df.
sum(stage_result_count)
nrow(df)

sum(extra_time_result_count)
nrow(df)

# Each row of the percentage tables should total approximately 100%.
rowSums(stage_result_percent)
rowSums(extra_time_result_percent)
```

`Result`, `Stage`, and `ExtraTime` should return `"factor"`. The missing-value checks should return zero. The totals from the two count tables should equal the total number of rows in `df`. Each row of the percentage tables should total approximately 100%. Small differences such as 99.9% or 100.1% may occur because the percentages were rounded.

## Results and Interpretation

```{r question-4-interpretation-values, include=FALSE}
# Find the column containing home-team wins.
home_result_column <- which(
  tolower(colnames(stage_result_percent)) ==
    "home team win"
)

# Find the Stage with the highest percentage of home-team wins.
highest_home_stage_row <- which.max(
  stage_result_percent[, home_result_column]
)

highest_home_stage <- rownames(
  stage_result_percent
)[highest_home_stage_row]

highest_home_stage_percent <- stage_result_percent[
  highest_home_stage_row,
  home_result_column
]

# Find the ExtraTime category with the highest percentage of home-team wins.
highest_home_extra_row <- which.max(
  extra_time_result_percent[, home_result_column]
)

highest_home_extra <- rownames(
  extra_time_result_percent
)[highest_home_extra_row]

highest_home_extra_percent <- extra_time_result_percent[
  highest_home_extra_row,
  home_result_column
]
```

Across all matches, the most common result was **`r most_common_result`**, which occurred in `r most_common_result_count` matches.

For competition stage, **`r highest_home_stage`** had the highest proportion of home-team wins, at `r highest_home_stage_percent`%. The percentage table and proportional bar chart show that the distribution of home-team wins, away-team wins, and draws was not identical across all competition stages. This suggests that `Stage` is associated with differences in match outcomes.

For extra time, category **`r highest_home_extra`** had the highest proportion of home-team wins, at `r highest_home_extra_percent`%. The percentage table and bar chart allow the result distribution for matches with different extra-time categories to be compared.

These findings show associations in the FIFA World Cup dataset. They do not prove that competition stage or extra time directly caused a particular match result.

\newpage

# Question 5: Git and GitHub

## Repository Link

https://github.com/yvette195/COMP1013-FIFA-WorldCup-Analysis

## Project Folder Structure

- `data/`: contains the original FIFA World Cup datasets.
- `scripts/`: contains the separate R scripts used for Questions 1–4.
- `report/`: contains the R Markdown source file and final PDF report.
- `figures/`: stores figures or screenshots used in the report.

## Commit History

_Insert the GitHub commit-history screenshot here._

## Explanation of Commits

_To be completed after the final commit history is available._
