---
title: "FIFA World Cup Data Analysis"
author: "Nguyễn Đào Trường Vy"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

library(tidyverse)
```

# Student Information

**Student Name:** Nguyễn Đào Trường Vy

**Student Number:** 22202382

**Subject Code:** COMP1013

**Subject Name:** Analytics Programming

**GitHub Repository:**  
https://github.com/yvette195/COMP1013-FIFA-WorldCup-Analysis

# Declaration

By including this statement, we the authors of this work, verify that:

- We hold a copy of this assignment that we can produce if the original is lost or damaged.
- We hereby certify that no part of this assignment/product has been copied from any other student’s work or from any other source except where due acknowledgement is made in the assignment.
- No part of this assignment/product has been written/produced for us by another person except where such collaboration has been authorised by the subject lecturer/tutor concerned.
- We are aware that this work may be reproduced and submitted to plagiarism detection software programs for the purpose of detecting possible plagiarism, which may retain a copy on its database for future plagiarism checking.
- We hereby certify that we have read and understand what the School of Computing, Engineering and Mathematics defines as minor and substantial breaches of misconduct as outlined in the learning guide for this unit.

\newpage

# Question 1: Data Inspection, Cleaning and Home Penalty Distribution

## Rationale

The four datasets were joined so that match, team, stadium and tournament information could be analysed together. Missing values written as `"?"` were changed to `NA`. The variables `Result`, `Stage`, `Country` and `ExtraTime` were changed to factors because they contain categories. `HomePenalty` and `AwayPenalty` were changed to numeric variables because they contain numbers of penalty goals.

A bar chart was selected because `HomePenalty` contains separate whole-number values. The height of each bar shows how many penalty-shootout matches had each home-team penalty score.

## Load and Join the Data

```{r question-1-load}
# Read the four CSV files.
# Report.Rmd is inside the report folder, so ../data goes back to the project folder.
matches <- read.csv("../data/Matches.csv")
stadiums <- read.csv("../data/Stadiums.csv")
teams <- read.csv("../data/Teams.csv")
tournaments <- read.csv("../data/Tournaments.csv")

# Change every "?" value to NA.
matches[matches == "?"] <- NA
stadiums[stadiums == "?"] <- NA
teams[teams == "?"] <- NA
tournaments[tournaments == "?"] <- NA
```

```{r question-1-join}
# Join team information for the home team and away team.
df <- matches %>%
  left_join(
    teams %>%
      rename(
        HomeTeamID = TeamID,
        HomeTeamName = TeamName,
        HomeTeamCode = TeamCode
      ),
    by = "HomeTeamID"
  ) %>%
  left_join(
    teams %>%
      rename(
        AwayTeamID = TeamID,
        AwayTeamName = TeamName,
        AwayTeamCode = TeamCode
      ),
    by = "AwayTeamID"
  ) %>%
  left_join(stadiums, by = "StadiumID") %>%
  left_join(tournaments, by = "TournamentID")
```

## Inspect and Clean the Data

```{r question-1-inspect}
# Display the first six rows.
head(df)

# Check the number of rows and columns.
dim(df)

# Check the variable types.
str(df)

# Display a summary of the main variables.
summary(
  df[c(
    "Result",
    "Stage",
    "Country",
    "ExtraTime",
    "PenaltyShootout",
    "HomePenalty",
    "AwayPenalty"
  )]
)
```

```{r question-1-clean}
# Convert categorical variables to factors.
df$Result <- as.factor(df$Result)
df$Stage <- as.factor(df$Stage)
df$Country <- as.factor(df$Country)
df$ExtraTime <- as.factor(df$ExtraTime)

# Convert penalty variables to numeric.
df$HomePenalty <- as.numeric(df$HomePenalty)
df$AwayPenalty <- as.numeric(df$AwayPenalty)

# Replace missing penalty values with zero.
df$HomePenalty[is.na(df$HomePenalty)] <- 0
df$AwayPenalty[is.na(df$AwayPenalty)] <- 0
```

## Code Testing

```{r question-1-testing}
# The number of rows should be unchanged after joining.
nrow(matches)
nrow(df)

# Check the classes after conversion.
class(df$Result)
class(df$Stage)
class(df$Country)
class(df$ExtraTime)
class(df$HomePenalty)
class(df$AwayPenalty)

# Check that no missing penalty values remain.
sum(is.na(df$HomePenalty))
sum(is.na(df$AwayPenalty))
```

The two row counts should be equal. The four categorical variables should return `"factor"`, the two penalty variables should return `"numeric"`, and both missing-value checks should return zero.

## Results and Interpretation

```{r question-1-results}
# Keep only matches involving a penalty shootout.
penalty_matches <- df %>%
  filter(PenaltyShootout == 1)

# Create a frequency table for HomePenalty.
penalty_table <- table(penalty_matches$HomePenalty)
penalty_table

# Find the most common HomePenalty value.
most_common_penalty <- names(penalty_table)[which.max(penalty_table)]
most_common_count <- max(penalty_table)
```

```{r question-1-plot, fig.width=8, fig.height=5}
ggplot(penalty_matches, aes(x = factor(HomePenalty))) +
  geom_bar() +
  labs(
    title = "Distribution of Home-Team Penalty Goals",
    x = "Home-team penalty goals",
    y = "Number of matches"
  ) +
  theme_minimal()
```

There were `r nrow(penalty_matches)` matches involving a penalty shootout. The most common value of `HomePenalty` was `r most_common_penalty`, and it occurred in `r most_common_count` matches. The taller bars show the penalty scores that occurred most often, while shorter bars represent less common scores.

\newpage

# Question 2: Distribution of Home Team Scores

## Rationale

`HomeTeamScore` is a numerical variable, so histograms were used to show its distribution. A bin width of one was selected because football scores are whole numbers. The distributions were compared first by `Stage`, then by `TournamentName`, and finally by the three required `AwayTeamScore` groups.

The `aggregate()` function was used to calculate simple group means and medians. This provides numerical information to support the patterns shown in the histograms.

## Distribution by Competition Stage

```{r question-2-stage-summary}
# Calculate the mean and median HomeTeamScore for each stage.
stage_mean <- aggregate(
  HomeTeamScore ~ Stage,
  mean,
  data = df
)

stage_median <- aggregate(
  HomeTeamScore ~ Stage,
  median,
  data = df
)

stage_mean
stage_median
```

```{r question-2-stage-plot, fig.width=9, fig.height=7}
ggplot(df, aes(x = HomeTeamScore)) +
  geom_histogram(binwidth = 1) +
  facet_wrap(~ Stage) +
  labs(
    title = "Home-Team Scores by Competition Stage",
    x = "Home-team goals",
    y = "Number of matches"
  ) +
  theme_minimal()
```

## Distribution by Tournament Name

```{r question-2-tournament-summary}
# Calculate the mean and median HomeTeamScore for each tournament.
tournament_mean <- aggregate(
  HomeTeamScore ~ TournamentName,
  mean,
  data = df
)

tournament_median <- aggregate(
  HomeTeamScore ~ TournamentName,
  median,
  data = df
)

tournament_mean
tournament_median
```

```{r question-2-tournament-plot, fig.width=10, fig.height=12}
ggplot(df, aes(x = HomeTeamScore)) +
  geom_histogram(binwidth = 1) +
  facet_wrap(~ TournamentName) +
  labs(
    title = "Home-Team Scores by Tournament",
    x = "Home-team goals",
    y = "Number of matches"
  ) +
  theme_minimal()
```

## Distribution by Away-Team Score Group

```{r question-2-away-groups}
# Create an empty variable for the away-team score groups.
df$AwayScoreGroup <- NA

# Assign each match to one of the three required groups.
df$AwayScoreGroup[df$AwayTeamScore <= 1] <- "0-1 goals"

df$AwayScoreGroup[
  df$AwayTeamScore >= 2 &
    df$AwayTeamScore <= 3
] <- "2-3 goals"

df$AwayScoreGroup[df$AwayTeamScore >= 4] <- "4 or more goals"

# Convert the new variable to a factor.
df$AwayScoreGroup <- as.factor(df$AwayScoreGroup)

# Check the number of matches in each group.
table(df$AwayScoreGroup)

# Calculate the mean and median HomeTeamScore for each group.
away_group_mean <- aggregate(
  HomeTeamScore ~ AwayScoreGroup,
  mean,
  data = df
)

away_group_median <- aggregate(
  HomeTeamScore ~ AwayScoreGroup,
  median,
  data = df
)

away_group_mean
away_group_median
```

```{r question-2-away-plot, fig.width=9, fig.height=5}
ggplot(df, aes(x = HomeTeamScore)) +
  geom_histogram(binwidth = 1) +
  facet_wrap(~ AwayScoreGroup) +
  labs(
    title = "Home-Team Scores by Away-Team Score Group",
    x = "Home-team goals",
    y = "Number of matches"
  ) +
  theme_minimal()
```

## Code Testing

```{r question-2-testing}
# Check the score variable types.
class(df$HomeTeamScore)
class(df$AwayTeamScore)

# Check for missing score values.
sum(is.na(df$HomeTeamScore))
sum(is.na(df$AwayTeamScore))

# Check the new group variable.
class(df$AwayScoreGroup)
table(df$AwayScoreGroup)
sum(is.na(df$AwayScoreGroup))

# Check that all matches are included in the three groups.
sum(table(df$AwayScoreGroup))
nrow(df)
```

`HomeTeamScore` and `AwayTeamScore` should be numeric variables; R may display their class as `integer` because football goals are whole numbers. The missing-value checks should return zero. `AwayScoreGroup` should be a factor, and the total number shown by its frequency table should equal the total number of rows in `df`.

## Results and Interpretation

```{r question-2-interpretation-values, include=FALSE}
# Find the stage with the highest and lowest mean HomeTeamScore.
highest_stage_row <- which.max(stage_mean$HomeTeamScore)
lowest_stage_row <- which.min(stage_mean$HomeTeamScore)

highest_stage <- stage_mean$Stage[highest_stage_row]
highest_stage_mean <- stage_mean$HomeTeamScore[highest_stage_row]

lowest_stage <- stage_mean$Stage[lowest_stage_row]
lowest_stage_mean <- stage_mean$HomeTeamScore[lowest_stage_row]

# Find the tournament with the highest and lowest mean HomeTeamScore.
highest_tournament_row <- which.max(tournament_mean$HomeTeamScore)
lowest_tournament_row <- which.min(tournament_mean$HomeTeamScore)

highest_tournament <- tournament_mean$TournamentName[highest_tournament_row]
highest_tournament_mean <- tournament_mean$HomeTeamScore[highest_tournament_row]

lowest_tournament <- tournament_mean$TournamentName[lowest_tournament_row]
lowest_tournament_mean <- tournament_mean$HomeTeamScore[lowest_tournament_row]

# Find the away-score group with the highest mean HomeTeamScore.
highest_group_row <- which.max(away_group_mean$HomeTeamScore)
highest_group <- away_group_mean$AwayScoreGroup[highest_group_row]
highest_group_mean <- away_group_mean$HomeTeamScore[highest_group_row]
```

The stage with the highest average home-team score was **`r highest_stage`**, with an average of `r round(highest_stage_mean, 2)` goals. The lowest average occurred in **`r lowest_stage`**, with `r round(lowest_stage_mean, 2)` goals. The stage histograms also show whether the scores were concentrated at low values or spread across a wider range.

The tournament with the highest average home-team score was **`r highest_tournament`**, with `r round(highest_tournament_mean, 2)` goals. The lowest tournament average occurred in **`r lowest_tournament`**, with `r round(lowest_tournament_mean, 2)` goals. The tournament histograms make it possible to compare how the score distributions changed between World Cup editions.

Among the three away-team score groups, **`r highest_group`** had the highest average home-team score at `r round(highest_group_mean, 2)` goals. This result describes a relationship in the dataset, but it does not prove that the away-team score caused the home-team score.

\newpage

# Question 3: Penalty Shootout Participation

## Rationale

_To be completed._

## Top Five Home Teams

```{r question-3-home-teams}
# Question 3 code will be added here.
```

## Top Five Away Teams

```{r question-3-away-teams}
# Question 3 code will be added here.
```

## Code Testing

_To be completed._

## Results and Interpretation

_To be completed._

\newpage

# Question 4: Factors Influencing Match Outcomes

## Selected Factors and Rationale

_To be completed._

## Analysis

```{r question-4-analysis}
# Question 4 code will be added here.
```

## Code Testing

_To be completed._

## Results and Interpretation

_To be completed._

\newpage

# Question 5: Git and GitHub

## Repository Link

https://github.com/yvette195/COMP1013-FIFA-WorldCup-Analysis

## Project Folder Structure

- `data/`: contains the original FIFA World Cup datasets.
- `scripts/`: contains the separate R scripts used for Questions 1–4.
- `report/`: contains the R Markdown source file and final PDF report.
- `figures/`: stores figures or screenshots used in the report.

## Commit History

_Insert the GitHub commit-history screenshot here._

## Explanation of Commits

_To be completed after the final commit history is available._
