---
title: "FIFA World Cup Data Analysis"
author: "Nguyễn Đào Trường Vy"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)

library(tidyverse)

# Allow the report to work whether it is run from the project root
# or knitted from inside the report folder.
project_root <- if (
  file.exists("COMP1013-FIFA-WorldCup-Analysis.Rproj")
) "." else ".."
```

# Student Information

**Student Name:** Nguyễn Đào Trường Vy

**Student Number:** 22202382

**Subject Code:** COMP1013

**Subject Name:** Analytics Programming

**GitHub Repository:**  
https://github.com/yvette195/COMP1013-FIFA-WorldCup-Analysis

# Declaration

By including this statement, we the authors of this work, verify that:

- We hold a copy of this assignment that we can produce if the original is lost or damaged.
- We hereby certify that no part of this assignment/product has been copied from any other student’s work or from any other source except where due acknowledgement is made in the assignment.
- No part of this assignment/product has been written/produced for us by another person except where such collaboration has been authorised by the subject lecturer/tutor concerned.
- We are aware that this work may be reproduced and submitted to plagiarism detection software programs for the purpose of detecting possible plagiarism, which may retain a copy on its database for future plagiarism checking.
- We hereby certify that we have read and understand what the School of Computing, Engineering and Mathematics defines as minor and substantial breaches of misconduct as outlined in the learning guide for this unit.

\newpage

# Question 1: Data Inspection, Cleaning and Home Penalty Distribution

## Rationale

The four datasets were combined using their identifier variables so that match, team, stadium and tournament information could be analysed in one data frame. Missing values recorded as `"?"` were converted into `NA` during data import so that R could recognise them correctly. `Result`, `Stage`, `Country` and `ExtraTime` were converted to factors because they represent categorical information. `HomePenalty` and `AwayPenalty` were converted to numeric variables because they contain counts of penalty goals.

A bar chart was selected to display `HomePenalty` because it is a discrete count variable with a small number of integer values. Each bar shows how frequently a particular number of home-team penalty goals occurred among matches involving a penalty shootout.

## Data Inspection and Preparation

```{r question-1-import}
# Import all four datasets.
# The na.strings argument converts every "?" into NA during import.
matches <- read.csv(
  file.path(project_root, "data", "Matches.csv"),
  na.strings = "?",
  stringsAsFactors = FALSE
)

stadiums <- read.csv(
  file.path(project_root, "data", "Stadiums.csv"),
  na.strings = "?",
  stringsAsFactors = FALSE
)

teams <- read.csv(
  file.path(project_root, "data", "Teams.csv"),
  na.strings = "?",
  stringsAsFactors = FALSE
)

tournaments <- read.csv(
  file.path(project_root, "data", "Tournaments.csv"),
  na.strings = "?",
  stringsAsFactors = FALSE
)
```

```{r question-1-join}
# Prepare a copy of the team data for home-team information.
home_teams <- teams %>%
  transmute(
    HomeTeamID = TeamID,
    HomeTeamName = TeamName,
    HomeTeamCode = TeamCode
  )

# Prepare a second copy for away-team information.
away_teams <- teams %>%
  transmute(
    AwayTeamID = TeamID,
    AwayTeamName = TeamName,
    AwayTeamCode = TeamCode
  )

# Join match, team, stadium and tournament information.
df <- matches %>%
  left_join(home_teams, by = "HomeTeamID") %>%
  left_join(away_teams, by = "AwayTeamID") %>%
  left_join(stadiums, by = "StadiumID") %>%
  left_join(tournaments, by = "TournamentID")
```

```{r question-1-inspection}
# Inspect the dimensions, variable names and variable types.
dim(df)
names(df)
str(df)

# Present the first six observations using the main analytical variables.
df %>%
  select(
    MatchID,
    TournamentName,
    Stage,
    HomeTeamName,
    AwayTeamName,
    HomeTeamScore,
    AwayTeamScore,
    PenaltyShootout
  ) %>%
  head(6) %>%
  knitr::kable(caption = "Sample of the combined FIFA World Cup dataset")

# Present a concise summary of the main variables used in Question 1.
summary(
  df[c(
    "Result",
    "Stage",
    "Country",
    "ExtraTime",
    "PenaltyShootout",
    "HomePenalty",
    "AwayPenalty"
  )]
)
```

```{r question-1-conversion}
# Convert categorical variables to factors.
# Convert penalty variables to numeric and replace missing values with zero.
df <- df %>%
  mutate(
    Result = as.factor(Result),
    Stage = as.factor(Stage),
    Country = as.factor(Country),
    ExtraTime = as.factor(ExtraTime),
    HomePenalty = replace_na(as.numeric(HomePenalty), 0),
    AwayPenalty = replace_na(as.numeric(AwayPenalty), 0)
  )
```

## Code Testing

```{r question-1-testing}
# Test that joining the datasets did not change the number of matches.
stopifnot(nrow(df) == nrow(matches))

# Check whether important joined fields contain missing values.
join_missing_values <- colSums(
  is.na(
    df[c(
      "HomeTeamName",
      "AwayTeamName",
      "StadiumName",
      "TournamentName"
    )]
  )
)

join_missing_values

# Confirm that the required variables have the intended data types.
converted_classes <- sapply(
  df[c(
    "Result",
    "Stage",
    "Country",
    "ExtraTime",
    "HomePenalty",
    "AwayPenalty"
  )],
  class
)

converted_classes

# Confirm that no missing values remain in the penalty variables.
penalty_missing_values <- colSums(
  is.na(df[c("HomePenalty", "AwayPenalty")])
)

penalty_missing_values

# Stop the report if any required conversion was unsuccessful.
stopifnot(
  is.factor(df$Result),
  is.factor(df$Stage),
  is.factor(df$Country),
  is.factor(df$ExtraTime),
  is.numeric(df$HomePenalty),
  is.numeric(df$AwayPenalty),
  !anyNA(df$HomePenalty),
  !anyNA(df$AwayPenalty)
)
```

## Results and Interpretation

```{r question-1-penalty-data}
# Retain only matches that involved a penalty shootout.
penalty_matches <- df %>%
  filter(PenaltyShootout == 1)

# Confirm that the filtered dataset contains observations.
stopifnot(nrow(penalty_matches) > 0)

# Count the frequency of each HomePenalty value.
penalty_distribution <- penalty_matches %>%
  count(HomePenalty, name = "NumberOfMatches") %>%
  arrange(HomePenalty)

# Calculate values used in the written interpretation.
most_common_penalty <- penalty_distribution %>%
  slice_max(NumberOfMatches, n = 1, with_ties = FALSE)

top_two_share <- penalty_distribution %>%
  slice_max(NumberOfMatches, n = 2, with_ties = FALSE) %>%
  summarise(
    Percentage = 100 * sum(NumberOfMatches) / nrow(penalty_matches)
  ) %>%
  pull(Percentage)

knitr::kable(
  penalty_distribution,
  caption = "Distribution of home-team penalty goals in penalty shootouts"
)
```

```{r question-1-plot, fig.width=8, fig.height=5}
ggplot(
  penalty_matches,
  aes(x = factor(HomePenalty))
) +
  geom_bar() +
  labs(
    title = "Distribution of Home-Team Penalty Goals",
    subtitle = "FIFA World Cup matches involving a penalty shootout",
    x = "Home-team penalty goals",
    y = "Number of matches"
  ) +
  theme_minimal()
```

There were `r nrow(penalty_matches)` matches involving a penalty shootout. `HomePenalty` ranged from `r min(penalty_matches$HomePenalty)` to `r max(penalty_matches$HomePenalty)` goals. The most frequently observed value was `r most_common_penalty$HomePenalty`, occurring in `r most_common_penalty$NumberOfMatches` matches. The two most frequent penalty totals together represented `r round(top_two_share, 1)`% of all penalty-shootout matches, showing that the distribution was concentrated around a small number of values rather than being evenly spread across the full range.

\newpage

# Question 2: Distribution of Home Team Scores

## Rationale

`HomeTeamScore` is a numerical count variable, so histograms are appropriate for showing its distribution. A bin width of one goal was selected because football scores occur as whole numbers. The bin boundary was set to `-0.5` so that each bar is centred on an integer score such as 0, 1, 2 or 3.

Separate faceted histograms are used for `Stage` and `TournamentName`. This is more readable than placing both variables in one very large grid. Summary tables containing the number of matches, mean, median, standard deviation and maximum score are also included to support the visual comparisons. For the second part, `AwayTeamScore` is grouped into `0–1 goals`, `2–3 goals`, and `4 or more goals`, as specified in the task.

## Analysis by Competition Stage

```{r question-2-stage-summary}
# Summarise the distribution of HomeTeamScore within each competition stage.
stage_score_summary <- df %>%
  group_by(Stage) %>%
  summarise(
    NumberOfMatches = n(),
    MeanHomeScore = mean(HomeTeamScore),
    MedianHomeScore = median(HomeTeamScore),
    SDHomeScore = sd(HomeTeamScore),
    MaximumHomeScore = max(HomeTeamScore),
    .groups = "drop"
  ) %>%
  arrange(desc(MeanHomeScore))

knitr::kable(
  stage_score_summary,
  digits = 2,
  caption = "Home-team score summary by competition stage"
)
```

```{r question-2-stage-histogram, fig.width=9, fig.height=7}
# Use one-goal bins and facet the histogram by Stage.
ggplot(df, aes(x = HomeTeamScore)) +
  geom_histogram(
    binwidth = 1,
    boundary = -0.5
  ) +
  facet_wrap(
    ~ Stage,
    ncol = 3,
    scales = "free_y"
  ) +
  scale_x_continuous(
    breaks = seq(
      0,
      max(df$HomeTeamScore),
      by = 1
    )
  ) +
  labs(
    title = "Distribution of Home-Team Scores by Competition Stage",
    x = "Home-team goals",
    y = "Number of matches"
  ) +
  theme_minimal()
```

## Analysis by Tournament Name

```{r question-2-tournament-summary}
# Summarise the distribution of HomeTeamScore within each tournament.
tournament_score_summary <- df %>%
  group_by(TournamentName) %>%
  summarise(
    NumberOfMatches = n(),
    MeanHomeScore = mean(HomeTeamScore),
    MedianHomeScore = median(HomeTeamScore),
    SDHomeScore = sd(HomeTeamScore),
    MaximumHomeScore = max(HomeTeamScore),
    .groups = "drop"
  ) %>%
  arrange(TournamentName)

knitr::kable(
  tournament_score_summary,
  digits = 2,
  caption = "Home-team score summary by tournament"
)
```

```{r question-2-tournament-histogram, fig.width=10, fig.height=13}
# Facet the histogram by TournamentName to compare World Cup editions.
ggplot(df, aes(x = HomeTeamScore)) +
  geom_histogram(
    binwidth = 1,
    boundary = -0.5
  ) +
  facet_wrap(
    ~ TournamentName,
    ncol = 4,
    scales = "free_y"
  ) +
  scale_x_continuous(
    breaks = seq(
      0,
      max(df$HomeTeamScore),
      by = 1
    )
  ) +
  labs(
    title = "Distribution of Home-Team Scores by Tournament",
    x = "Home-team goals",
    y = "Number of matches"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 7)
  )
```

## Analysis by Away-Team Score Group

```{r question-2-away-score-groups}
# Create the three AwayTeamScore groups required by the task.
df <- df %>%
  mutate(
    AwayScoreGroup = case_when(
      AwayTeamScore <= 1 ~ "0-1 goals",
      AwayTeamScore <= 3 ~ "2-3 goals",
      AwayTeamScore >= 4 ~ "4 or more goals"
    ),
    AwayScoreGroup = factor(
      AwayScoreGroup,
      levels = c(
        "0-1 goals",
        "2-3 goals",
        "4 or more goals"
      )
    )
  )

# Summarise HomeTeamScore within each away-team score group.
away_group_summary <- df %>%
  group_by(AwayScoreGroup) %>%
  summarise(
    NumberOfMatches = n(),
    MeanHomeScore = mean(HomeTeamScore),
    MedianHomeScore = median(HomeTeamScore),
    SDHomeScore = sd(HomeTeamScore),
    MaximumHomeScore = max(HomeTeamScore),
    .groups = "drop"
  )

# Identify the most common HomeTeamScore in each group.
away_group_modes <- df %>%
  count(
    AwayScoreGroup,
    HomeTeamScore,
    name = "NumberOfMatches"
  ) %>%
  group_by(AwayScoreGroup) %>%
  slice_max(
    NumberOfMatches,
    n = 1,
    with_ties = FALSE
  ) %>%
  ungroup()

knitr::kable(
  away_group_summary,
  digits = 2,
  caption = "Home-team score summary by away-team score group"
)

knitr::kable(
  away_group_modes,
  caption = "Most common home-team score within each away-team score group"
)
```

```{r question-2-away-group-histogram, fig.width=9, fig.height=5}
# Compare HomeTeamScore distributions across the three AwayTeamScore groups.
ggplot(df, aes(x = HomeTeamScore)) +
  geom_histogram(
    binwidth = 1,
    boundary = -0.5
  ) +
  facet_wrap(
    ~ AwayScoreGroup,
    nrow = 1,
    scales = "free_y"
  ) +
  scale_x_continuous(
    breaks = seq(
      0,
      max(df$HomeTeamScore),
      by = 1
    )
  ) +
  labs(
    title = "Distribution of Home-Team Scores by Away-Team Score Group",
    x = "Home-team goals",
    y = "Number of matches"
  ) +
  theme_minimal()
```

## Code Testing

```{r question-2-testing}
# Confirm that both score variables are numeric, complete and non-negative.
stopifnot(
  is.numeric(df$HomeTeamScore),
  is.numeric(df$AwayTeamScore),
  !anyNA(df$HomeTeamScore),
  !anyNA(df$AwayTeamScore),
  all(df$HomeTeamScore >= 0),
  all(df$AwayTeamScore >= 0)
)

# Confirm that every match was assigned to exactly one away-score group.
away_group_counts <- table(
  df$AwayScoreGroup,
  useNA = "ifany"
)

away_group_counts

stopifnot(
  !anyNA(df$AwayScoreGroup),
  sum(away_group_counts) == nrow(df),
  all(levels(df$AwayScoreGroup) == c(
    "0-1 goals",
    "2-3 goals",
    "4 or more goals"
  )),
  sum(stage_score_summary$NumberOfMatches) == nrow(df),
  sum(tournament_score_summary$NumberOfMatches) == nrow(df),
  sum(away_group_summary$NumberOfMatches) == nrow(df)
)
```

## Results and Interpretation

```{r question-2-interpretation-values, include=FALSE}
# Calculate values used in the written interpretation.
highest_stage <- stage_score_summary %>%
  slice_max(
    MeanHomeScore,
    n = 1,
    with_ties = FALSE
  )

lowest_stage <- stage_score_summary %>%
  slice_min(
    MeanHomeScore,
    n = 1,
    with_ties = FALSE
  )

highest_tournament <- tournament_score_summary %>%
  slice_max(
    MeanHomeScore,
    n = 1,
    with_ties = FALSE
  )

lowest_tournament <- tournament_score_summary %>%
  slice_min(
    MeanHomeScore,
    n = 1,
    with_ties = FALSE
  )

highest_away_group <- away_group_summary %>%
  slice_max(
    MeanHomeScore,
    n = 1,
    with_ties = FALSE
  )

home_score_zero_to_three <- 100 * mean(
  df$HomeTeamScore >= 0 &
    df$HomeTeamScore <= 3
)
```

Across competition stages, **`r highest_stage$Stage`** had the highest average home-team score at `r round(highest_stage$MeanHomeScore, 2)` goals, while **`r lowest_stage$Stage`** had the lowest average at `r round(lowest_stage$MeanHomeScore, 2)` goals. The faceted histograms show how the centre, spread and frequency of high-scoring matches differ across stages. The free vertical scales are used because the number of matches varies substantially between stages.

Across tournament editions, **`r highest_tournament$TournamentName`** recorded the highest average home-team score at `r round(highest_tournament$MeanHomeScore, 2)` goals, whereas **`r lowest_tournament$TournamentName`** recorded the lowest average at `r round(lowest_tournament$MeanHomeScore, 2)` goals. Overall, `r round(home_score_zero_to_three, 1)`% of all home-team scores were between 0 and 3 goals, indicating that most matches were concentrated in the lower scoring range.

For the away-team score groups, **`r highest_away_group$AwayScoreGroup`** had the highest average home-team score at `r round(highest_away_group$MeanHomeScore, 2)` goals. The grouped histograms and mode table show whether higher away-team scoring is associated with a shift in the home-team score distribution. These results describe an association in the observed matches and should not be interpreted as evidence that the away-team score causes the home-team score.

\newpage

# Question 3: Penalty Shootout Participation

## Rationale

_To be completed._

## Top Five Home Teams

```{r question-3-home-teams}
# Question 3 code will be added here.
```

## Top Five Away Teams

```{r question-3-away-teams}
# Question 3 code will be added here.
```

## Code Testing

_To be completed._

## Results and Interpretation

_To be completed._

\newpage

# Question 4: Factors Influencing Match Outcomes

## Selected Factors and Rationale

_To be completed._

## Analysis

```{r question-4-analysis}
# Question 4 code will be added here.
```

## Code Testing

_To be completed._

## Results and Interpretation

_To be completed._

\newpage

# Question 5: Git and GitHub

## Repository Link

https://github.com/yvette195/COMP1013-FIFA-WorldCup-Analysis

## Project Folder Structure

- `data/`: contains the original FIFA World Cup datasets.
- `scripts/`: contains the separate R scripts used for Questions 1–4.
- `report/`: contains the R Markdown source file and final PDF report.
- `figures/`: stores figures or screenshots used in the report.

## Commit History

_Insert the GitHub commit-history screenshot here._

## Explanation of Commits

_To be completed after the final commit history is available._
