---
title: "FIFA World Cup Data Analysis"
author: "Nguyễn Đào Trường Vy"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)

library(tidyverse)

# Allow the report to work whether it is run from the project root
# or knitted from inside the report folder.
project_root <- if (
  file.exists("COMP1013-FIFA-WorldCup-Analysis.Rproj")
) "." else ".."
```

# Student Information

**Student Name:** Nguyễn Đào Trường Vy

**Student Number:** 22202382

**Subject Code:** COMP1013

**Subject Name:** Analytics Programming

**GitHub Repository:**  
https://github.com/yvette195/COMP1013-FIFA-WorldCup-Analysis

# Declaration

By including this statement, we the authors of this work, verify that:

- We hold a copy of this assignment that we can produce if the original is lost or damaged.
- We hereby certify that no part of this assignment/product has been copied from any other student’s work or from any other source except where due acknowledgement is made in the assignment.
- No part of this assignment/product has been written/produced for us by another person except where such collaboration has been authorised by the subject lecturer/tutor concerned.
- We are aware that this work may be reproduced and submitted to plagiarism detection software programs for the purpose of detecting possible plagiarism, which may retain a copy on its database for future plagiarism checking.
- We hereby certify that we have read and understand what the School of Computing, Engineering and Mathematics defines as minor and substantial breaches of misconduct as outlined in the learning guide for this unit.

\newpage

# Question 1: Data Inspection, Cleaning and Home Penalty Distribution

## Rationale

The four datasets were combined using their identifier variables so that match, team, stadium and tournament information could be analysed in one data frame. Missing values recorded as `"?"` were converted into `NA` during data import so that R could recognise them correctly. `Result`, `Stage`, `Country` and `ExtraTime` were converted to factors because they represent categorical information. `HomePenalty` and `AwayPenalty` were converted to numeric variables because they contain counts of penalty goals.

A bar chart was selected to display `HomePenalty` because it is a discrete count variable with a small number of integer values. Each bar shows how frequently a particular number of home-team penalty goals occurred among matches involving a penalty shootout.

## Data Inspection and Preparation

```{r question-1-import}
# Import all four datasets.
# The na.strings argument converts every "?" into NA during import.
matches <- read.csv(
  file.path(project_root, "data", "Matches.csv"),
  na.strings = "?",
  stringsAsFactors = FALSE
)

stadiums <- read.csv(
  file.path(project_root, "data", "Stadiums.csv"),
  na.strings = "?",
  stringsAsFactors = FALSE
)

teams <- read.csv(
  file.path(project_root, "data", "Teams.csv"),
  na.strings = "?",
  stringsAsFactors = FALSE
)

tournaments <- read.csv(
  file.path(project_root, "data", "Tournaments.csv"),
  na.strings = "?",
  stringsAsFactors = FALSE
)
```

```{r question-1-join}
# Prepare a copy of the team data for home-team information.
home_teams <- teams %>%
  transmute(
    HomeTeamID = TeamID,
    HomeTeamName = TeamName,
    HomeTeamCode = TeamCode
  )

# Prepare a second copy for away-team information.
away_teams <- teams %>%
  transmute(
    AwayTeamID = TeamID,
    AwayTeamName = TeamName,
    AwayTeamCode = TeamCode
  )

# Join match, team, stadium and tournament information.
df <- matches %>%
  left_join(home_teams, by = "HomeTeamID") %>%
  left_join(away_teams, by = "AwayTeamID") %>%
  left_join(stadiums, by = "StadiumID") %>%
  left_join(tournaments, by = "TournamentID")
```

```{r question-1-inspection}
# Inspect the dimensions, variable names and variable types.
dim(df)
names(df)
str(df)

# Present a concise summary of the main variables used in Question 1.
summary(
  df[c(
    "Result",
    "Stage",
    "Country",
    "ExtraTime",
    "PenaltyShootout",
    "HomePenalty",
    "AwayPenalty"
  )]
)
```

```{r question-1-conversion}
# Convert categorical variables to factors.
# Convert penalty variables to numeric and replace missing values with zero.
df <- df %>%
  mutate(
    Result = as.factor(Result),
    Stage = as.factor(Stage),
    Country = as.factor(Country),
    ExtraTime = as.factor(ExtraTime),
    HomePenalty = replace_na(as.numeric(HomePenalty), 0),
    AwayPenalty = replace_na(as.numeric(AwayPenalty), 0)
  )
```

## Code Testing

```{r question-1-testing}
# Test that joining the datasets did not change the number of matches.
stopifnot(nrow(df) == nrow(matches))

# Check whether important joined fields contain missing values.
join_missing_values <- colSums(
  is.na(
    df[c(
      "HomeTeamName",
      "AwayTeamName",
      "StadiumName",
      "TournamentName"
    )]
  )
)

join_missing_values

# Confirm that the required variables have the intended data types.
converted_classes <- sapply(
  df[c(
    "Result",
    "Stage",
    "Country",
    "ExtraTime",
    "HomePenalty",
    "AwayPenalty"
  )],
  class
)

converted_classes

# Confirm that no missing values remain in the penalty variables.
penalty_missing_values <- colSums(
  is.na(df[c("HomePenalty", "AwayPenalty")])
)

penalty_missing_values

# Stop the report if any required conversion was unsuccessful.
stopifnot(
  is.factor(df$Result),
  is.factor(df$Stage),
  is.factor(df$Country),
  is.factor(df$ExtraTime),
  is.numeric(df$HomePenalty),
  is.numeric(df$AwayPenalty),
  !anyNA(df$HomePenalty),
  !anyNA(df$AwayPenalty)
)
```

## Results and Interpretation

```{r question-1-penalty-data}
# Retain only matches that involved a penalty shootout.
penalty_matches <- df %>%
  filter(PenaltyShootout == 1)

# Confirm that the filtered dataset contains observations.
stopifnot(nrow(penalty_matches) > 0)

# Count the frequency of each HomePenalty value.
penalty_distribution <- penalty_matches %>%
  count(HomePenalty, name = "NumberOfMatches") %>%
  arrange(HomePenalty)

knitr::kable(
  penalty_distribution,
  caption = "Distribution of home-team penalty goals in penalty shootouts"
)
```

```{r question-1-plot, fig.width=8, fig.height=5}
ggplot(
  penalty_matches,
  aes(x = factor(HomePenalty))
) +
  geom_bar() +
  labs(
    title = "Distribution of Home-Team Penalty Goals",
    subtitle = "FIFA World Cup matches involving a penalty shootout",
    x = "Home-team penalty goals",
    y = "Number of matches"
  ) +
  theme_minimal()
```

There were `r nrow(penalty_matches)` matches involving a penalty shootout. The most frequently observed value of `HomePenalty` was `r penalty_distribution$HomePenalty[which.max(penalty_distribution$NumberOfMatches)]`, which occurred in `r max(penalty_distribution$NumberOfMatches)` matches. The chart shows whether the penalty-goal totals are concentrated around a small group of values or spread across a wider range. Values with shorter bars occurred less frequently.

\newpage

# Question 2: Distribution of Home Team Scores

## Rationale

_To be completed._

## Analysis by Competition Stage and Tournament

```{r question-2-stage-tournament}
# Question 2 code will be added here.
```

## Analysis by Away Team Score Groups

```{r question-2-away-score-groups}
# Question 2 code will be added here.
```

## Code Testing

_To be completed._

## Results and Interpretation

_To be completed._

\newpage

# Question 3: Penalty Shootout Participation

## Rationale

_To be completed._

## Top Five Home Teams

```{r question-3-home-teams}
# Question 3 code will be added here.
```

## Top Five Away Teams

```{r question-3-away-teams}
# Question 3 code will be added here.
```

## Code Testing

_To be completed._

## Results and Interpretation

_To be completed._

\newpage

# Question 4: Factors Influencing Match Outcomes

## Selected Factors and Rationale

_To be completed._

## Analysis

```{r question-4-analysis}
# Question 4 code will be added here.
```

## Code Testing

_To be completed._

## Results and Interpretation

_To be completed._

\newpage

# Question 5: Git and GitHub

## Repository Link

https://github.com/yvette195/COMP1013-FIFA-WorldCup-Analysis

## Project Folder Structure

- `data/`: contains the original FIFA World Cup datasets.
- `scripts/`: contains the separate R scripts used for Questions 1–4.
- `report/`: contains the R Markdown source file and final PDF report.
- `figures/`: stores figures or screenshots used in the report.

## Commit History

_Insert the GitHub commit-history screenshot here._

## Explanation of Commits

_To be completed after the final commit history is available._
